---
title: "Homework 1"
author: "Han Nguyen - TXN200004"
date: "09/02/2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1

## a)

```{r, problem_1a}
a <- round(8 + 9 - 7 / 3 ^ 0.3, 2)
```

**Answer:** $8 + 9 - \frac{7}{3^{0.3}}$ = ```r a```

## b)

```{r, problem_1b}
b <- round(log(sqrt((15 + 6) / (14 + 2)), base = 2), 2)
```

**Answer:** $\log_2 \left( \sqrt{\frac{15 + 6}{14 + 2}} \right)$ = ```r b```


## c)

```{r, problem_1c}
c <- round(((11 + sin(pi/4)) / (factorial(3) + abs(-10))) ^ 2, 2)
```

**Answer:** $\left( \frac{11 + \sin\left(\frac{\pi}{4}\right)}{3! + |-10|} \right)^2$ = ```r c```

## d)

```{r, problem_1d}
d <- round(6 + 5 - 4/3^2, 2)
```

**Answer:** $6 + 5 - \frac{4}{3^2}$ = ```r d```

## e)

```{r, problem_1e}
e <- round(exp( sqrt( (14 + 13) / (12 + 11) ) ), 2)
```

**Answer:** $e^{\sqrt{\frac{14 + 13}{12 + 11}}}$ = ```r e```

## f)

```{r, problem_1f}
f <- round(((11 + factorial(12)) / (factorial(13) + 14)) ^ 2, 2)
```

**Answer:** $\left( \frac{11 + 12!}{13! + 14} \right)^2$ = ```r f```

# Problem 2

## a)

```{r, problem_2a}
RF <- c(2.60, 3.05, 3.74, 3.48, 5.49, 4.25, 2.57, 2.18, 3.14, 4.82, 3.28, 3.01)
```

**Answer:** RF = ```r RF```

## b)

```{r, problem_2b}
names(RF) <- month.abb
```

**Answer:** The names of the RF vector are assigned to the abbreviated month names: ```r names(RF)```

## c)

```{r, problem_2c}
avg_RF <- round(mean(RF), 2)
```

**Answer:** The average RF value is ```r avg_RF```

## d)

```{r, problem_2d}
min_month <- names(which.min(RF))
max_month <- names(which.max(RF))
```

**Answer:** $\\$
- Month with the minimum RF is ```r min_month```. $\\$
- Month with the maximum RF is ```r max_month```.

# Problem 3

## a)

```{r, problem_3a}
H2 <- c(2700, 2600, 3050, 2900, 3000, 2500, 2600, 3000, 2800, 3200, 2800, 3400)
```
**Answer:** H2 = ```r H2```

## b)

```{r, problem_3b}
names(H2) <- month.abb
```

**Answer:** The names of the H2 vector are assigned to the abbreviated month names ```r names(H2)```

## c)

```{r, problem_3c}
total <- sum(H2)
```

**Answer:** Total sales of Hummer H2 in 2002 = ```r format(total, scientific = FALSE, big.mark = ",")```

## d)

```{r, problem_3d}
diff <- diff(H2)

max_increase <- names(which.max(diff))
max_decrease <- names(which.min(diff))
```
**Answer:** $\\$
Month with greatest increase: ```r max_increase```.$\\$
Month with greatest decrease ```r max_decrease```.

# Problem 4

## a)

```{r, problem_4a}
x <- c(1, -2, 3, -4, 5, 100)
y <- x * -1
y[y > 0]
```

## b)

```{r, problem_4b}
# create a sequence from 1 to 50
z <- seq(1, 50)

# test whether an observation is even
even <- z %% 2 == 0

# subset z by the test above
z <- z[even]
```

## c)

```{r, problem_4c}
mean <- function(x) {
  sum(x) / length(x)
}
```

# Problem 5

```{r, problem_5}
PrintSquare <- function() {
  for (i in 1:1000) {
    square_root <- sqrt(i)
    if (square_root == as.integer(square_root)) {
      print(i)
    }
  }
}

#Testing
PrintSquare()
```

# Problem 6

```{r, problem_6}
TwinPrime <- function(n) {
  # Function to check if a number is prime
  IsPrime <- function(x) {
    # Numbers <= 1 are not prime
    if (x <= 1) {
      return(FALSE)
    }
    
    # 2 and 3 are prime
    if (x == 2 || x == 3) {
      return(TRUE)
    }

    # Check up to square root of x
    # Skip 4 because it's not prime
    i <- 5
    for (i in 2:floor(sqrt(x))) {
      if (x %% i == 0) {
        return(FALSE)
      }
    }
      return(TRUE)
  }
  
  # Constructing a primes vector from 1 -> n
  primes <- c()
  for (i in 2:n) {
    if (IsPrime(i)) {
      # new vector and add new prime in
      primes <- c(primes, i)
    }
  }
  
  # Find and count all twin pairs
  total_pair <- 0
  for (i in 1:(length(primes) - 1)) {
    if (primes[i + 1] - primes[i] == 2) {
      total_pair <- total_pair + 1
    }
  }
  
  return(total_pair)
}

# Testing
TwinPrime(100)
```


