---
title: "Homework 7"
author: "Han Nguyen - TXN200004"
date: "11/13/2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

```{r, question1_data}
# Data: Cola can fill amounts (ounces)
fill <- c(15.997, 16.005, 15.981, 15.954, 15.986, 16.021, 15.985, 16.001, 16.018, 16.056)
```

**Step 1: State the null and alternative hypotheses**

- $H_0$: $\mu = 16.00$ (The mean fill of cola cans is 16.00 ounces)
- $H_a$: $\mu < 16.00$ (The mean fill of cola cans is less than 16.00 ounces)

This is a one-sided (left-tailed) test.

**Step 2: Choose the significance level**

$\alpha = 0.05$

**Step 3: Determine the appropriate test statistic**

Since we have a small sample (n = 10) and we don't know the population standard deviation, we use a one-sample t-test with the test statistic:

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

where $\bar{x}$ is the sample mean, $\mu_0 = 16.00$ is the hypothesized mean, $s$ is the sample standard deviation, and $n = 10$ is the sample size.

**Step 4: Formulate the decision rule**

With $\alpha = 0.05$ and degrees of freedom $df = n - 1 = 9$, we will reject $H_0$ if:

- $t < -t_{\alpha, df}$ (critical value approach), or
- p-value $< \alpha$ (p-value approach)

```{r, question1_critical_value}
# Critical value for one-sided test
alpha <- 0.05
df <- length(fill) - 1
t_critical <- qt(alpha, df)
```

The critical value is $t_{0.05, 9} =$ `r round(t_critical, 4)`.

**Step 5: Collect data and calculate the test statistic**

```{r, question1_calculations}
# Sample statistics
n <- length(fill)
x_bar <- mean(fill)
s <- sd(fill)
mu_0 <- 16.00

# Calculate test statistic
t_stat <- (x_bar - mu_0) / (s / sqrt(n))

# Calculate p-value (one-sided, left-tailed)
p_value <- pt(t_stat, df)
```

Sample mean: $\bar{x} =$ `r round(x_bar, 4)` ounces

Sample standard deviation: $s =$ `r round(s, 4)` ounces

Test statistic: $t =$ `r round(t_stat, 4)`

**Step 6: Make a decision**

```{r, question1_decision}
# Decision based on critical value
reject_H0_critical <- t_stat < t_critical

# Decision based on p-value
reject_H0_pvalue <- p_value < alpha
```

P-value = `r round(p_value, 4)`

Since $t =$ `r round(t_stat, 4)` is `r ifelse(reject_H0_critical, "less than", "greater than")` the critical value of `r round(t_critical, 4)`, and the p-value (`r round(p_value, 4)`) is `r ifelse(reject_H0_pvalue, "less than", "greater than")` $\alpha = 0.05$, we `r ifelse(reject_H0_pvalue, "reject", "fail to reject")` $H_0$.

**Step 7: Draw a conclusion**

Based on the sample of 10 cola cans, we `r ifelse(reject_H0_pvalue, "reject", "fail to reject")` the null hypothesis at the 0.05 significance level. There `r ifelse(reject_H0_pvalue, "is", "is not")` sufficient statistical evidence to conclude that the mean fill of cola cans is less than 16.00 ounces. `r ifelse(reject_H0_pvalue, "Mark's concern appears to be justified - the cans may contain less than the advertised 16 ounces.", "Mark's data does not support the claim that the cans contain less than 16 ounces.")`


# Question 2

```{r, question2}
# If I wrote code that would be used in all following letters, it would go here
```

# Question 3

```{r, question3}
# If I wrote code that would be used in all following letters, it would go here
```

# Question 4

```{r, question4}
# If I wrote code that would be used in all following letters, it would go here
```

# Question 5

```{r, question_5}
# This is where my code for this question goes
```

# Question 6

```{r, question_6}
# This is where my code for this question goes
```

# Question 7

```{r, question_7}
# This is where my code for this question goes
```

# Question 8

```{r, question_8}
# This is where my code for this question goes
```
